### 2021-12-12 documents/sections/verbs.tex
more finetuning

\subsection{Conservative verbs in comparison}
\label{sec:verbs}
In \cref{sec:extensions}, six incomplete extensions of personal prefixes into \gl{1}\gl{s_a_} territory were presented, and the verbs resistant to these innovations were identified.
The set of unaffected verbs is rather small in most cases, and (proto-)languages show a considerable degree of overlap in what verbs are unaffected.
Here,  resistant verbs are investigated from a comparative perspective:
\rc{ka[ti]} \qu{to say} \pcref{sec:say}, \rc{ɨtə[mə]} \qu{to go} \pcref{sec:go}, both roots of the copula \rc{eti}/\obj{a[p]} \pcref{sec:be}, \rc{(ət)jəpɨ} \qu{to come} \pcref{sec:come},  \rc{ɨpɨtə} \qu{to go down} \pcref{sec:godown}, and  \cref{sec:bathe} investigates \rc{e-pɨ} \qu{to bathe}, of which the \PPek reflex \rc{ipɨ} resisted the extension of \rc{k-}.
The \obj{e}-initial verbs not affected by the extension of \obj{k-} in \akuriyo \pcref{sec:akuriyo} will not be discussed here, as they are a large and phonologically coherent group.

\subsubsection{\rc{ka[ti]} \qu{to say}}
\label{sec:say}
Most reflexes of this verb are simply \obj{ka}, but a fleeting syllable \rc{ti} is reconstructed by \textcite{gildea2007greenberg}, best visible in the imperative forms of some languages.% \exref{kati}.
%
%\ex<kati> \apalai\\
%\obj{kaʃi-ko} \qu{say!}\\
%\parencite[35]{koehn1986apalai}
%\xe
%
\cref{tab:say} shows a comparison of the longest attested forms for each language.%
\footnote{Cognate segments in \crefrange{tab:say}{tab:bathe} were aligned automatically with LingPy \parencite{lingpy268}, for easier recognition of correspondences.}
%
\input{floats/say.tex}%
%
This verb was not affected by any of the extensions found in \PPek, \PWai, \PTir, \akuriyo, or \carijo (\crefrange{sec:pekodian}{sec:carijo}).
The first person form of its \yukpa reflex \obj{ka} is unattested.

As mentioned in \cref{sec:waiwaian}, \textcite{hixkaryanaderby1985} analyzes this verb as transitive in \hixka.
This analytical choice is not only motivated by a desire to avoid an idiosyncratic intransitive first person prefix \obj{ɨ-} instead of regular \obj{kɨ-}.
\hixka \obj{ka} also shows the complementary distribution of third person  \obj{n-} and preceding objects typical of transitive verbs in Cariban \parencite[60--81]{gildea1998}.
Due to the semantics of \qu{to say}, these objects are either ideophones or direct speech \exref{hixsay}.

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}).
It does not occur in \exref{hixsay.ordpl-35}, because \qu{they said} is preceded by direct speech.
Looking beyond \hixka, at least the \trio cognate shows the same pattern, albeit inconsistently so \parencite[267]{triocarlin2004}.

Derivational suffixes also point to \rc{ka[ti]} \qu{to say} being transitive:
\trio \obj{ka} is characterized as the only intransitive verb being able to take the causative suffix \obj{-po} and the agentive nominalizer \obj{-ne} \parencite[263, 169]{triomeira1999}.
The exceptionality of \obj{ka} \qu{to say} taking \obj{-po} \qu{\gl{caus}.\gl{tr}} has also been noted for \kalina \parencite[82]{courtz2008carib} and \wayana \parencite[258]{wayanatavares2005}.
The agent nominalizer \rc{-ne} gave rise to the \panare gnomic verbal suffix \obj{-ɲe} on transitive verbs \parencite[184--185]{gildea1998}.
The occurrence of \obj{-ɲe} on \obj{ka} likely led \textcite[214]{panarepayne2013} to categorize it as transitive, contrasting with the intransitive analysis by \textcite[102]{mattei1994diccionario}.
Finally, reflexes of the causativizer \rc{-metipo}, usually restricted to transitive verbs \parencite{gildea2015valency}, are found with \obj{ka} in \apalai \parencite[51]{koehn1986apalai} and \waiwai \parencite[52]{waiwaihawkins1998}.

The categorization of \qu{to say} as an intransitive verb is supported primarily by its person prefixes.
\kalina offers a minimal pair between transitive \obj{ka} \qu{to remove} and intransitive \obj{ka} \qu{to say}, \obj{sikai} \qu{I took it away} vs \obj{wɨkai} \qu{I said} \parencite[288, 45]{courtz2008carib}.
Similarly, \qu{to say} has a reflex of \gl{1}\gl{s_a_} \rc{w-} in Pekodian languages \pcref{sec:pekodian}, rather than \gl{1}>\gl{3} \obj{s-} (\bakairi) or \rc{ini-} (\PXin).
Additionally, languages which differentiate direct prefixes from \gl{s_a_} prefixes by the presence of \obj{i} \parencite[495]{meira2010origin} show \obj{ɨ} instead of \obj{i} for this verb, see \akuriyo in \exref{sayintr.aku-154}, as well as \textcites[294]{triomeira1999}[195]{wayanatavares2005}[288]{ikpengpacheco2001}[150]{alves2017arara}[168]{hoff1968carib} for cognate forms in other such languages.
Finally, the \gl{s_a_} class marker \obj{w-} occurs on nominalizations in \kalina \exref{sayintr.kar-84}, and it is probably reflected as vowel length in the \trio \parencite[333]{triomeira1999} and \wayana \parencite[196]{wayanatavares2005} participles.

@@

\subsubsection{\rc{ɨtə[mə]} \qu{to go}}
\label{sec:go}
This verb is reconstructed by \textcite{gildea2007greenberg} as \rc{tə[mə]}, like \rc{ka[ti]} \qu{to say} with a fleeting second syllable.
It is true that many reflexes are clearly \obj{t}-initial, for example \hixka \obj{ntoje} \qu{he went} \parencite[27]{hixkaryanaderby1985}, \trio \obj{təkə} \qu{go!} \parencite[246]{triomeira1999}, or \wayana \obj{kuptəm} \qu{we went} \parencite[195]{wayanatavares2005}.
However, once one considers all forms of the various reflexes of this verb \pcref{tab:go}, an initial vowel \rc{ɨ} must clearly be reconstructed -- in contrast to unambiguously C-initial \rc{ka[ti]} \qu{to say}.%
\footnote{As indicated by the brackets in \cref{tab:go}, there are many languages where the initial vowel is only present in some forms.
Also, the prefix-verb boundary in many inflected forms like e.g.\ \trio \obj{wɨtənne} or \arara \obj{wɨdolɨ} \qu{I went} \parencites[43]{triomeira1999}[153]{alves2017arara} is ambiguous, since an epenthetic \obj{ɨ} breaks up potential CC clusters.
Still, when one considers unambiguous forms, the contrast with \rc{ka[ti]} becomes very clear.}
This verb was not affected by the any of the extensions in \cref{sec:extensions}.

@@

\subsubsection{\rc{eti} and \rc{a[p]} \qu{to be}}
\label{sec:be}
For a comprehensive comparative overview of these two roots, readers are referred to \textcite[375--382]{gildea2018reconstructing}; they will not be discussed in detail here.
\rc{a[p]} is the original copula and can be reconstructed as already having various irregularities in \PC.
\rc{eti} is reconstructed by \textcites{meira2009property}{gildea2018reconstructing} as originally meaning \qu{to dwell, live}, but serving as a copula already in \PC.\footnote{Such a stative, locative source is also suggested by the existence of \obj{it͡ʃi} \qu{to lie down} in \arara \parencite[196]{alves2017arara}.}
Various modern languages use reflexes of these two roots in a suppletive manner, conditioned by person and\slash{}or \gl{tam}.
Both roots preserved \gl{1}\gl{s_a_} \rc{w-} in \PPek, \PWai, and \PTir (\crefrange{sec:pekodian}{sec:taranoan}).
\akuriyo \obj{a} was not affected by the extension of \obj{k-} \pcref{sec:akuriyo}, while \obj{eʔi} is not attested in a first-person form. 
\carijo innovated \obj{j-}, but only in the reflex of \rc{eti} \exref{mivida-12}; the \obj{a} root preserves \obj{w-} \pcref{sec:carijo}.
\yukpa introduced \obj{j-} to the reflexes of both \rc{a[p]} and \rc{eti}, which are preserved as encliticized auxiliaries in certain constructions \exref{yukcop}.

@@

\ex<yukcop> \yukpa \parencite[143--144]{meira2006syntactic}\\
\begin{tabular}[t]{@{}lll@{}}
	& \gl{npst} & \gl{pst}\\
	\gl{1} & \obj{=j-a(-s)}&\obj{=j-e}\\
	\gl{2} & \obj{=mak(o)}&\obj{=m-e}\\
	\gl{3} & \obj{=mak(o)}&\obj{=n-e}\\
\end{tabular}
\xe

\subsubsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}

@@

This verb is reconstructed as \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all attested reflexes \pcref{tab:come} points to a more complex story.
Crucially, the majority do not reflect the \rc{ət} part of their reconstruction, and many forms are ostensibly reflexes of \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ}, rather than \rc{epɨ}.
A unifying account of all these forms is achieved by reconstructing a \PC form \rc{(ət-)jəpɨ}, morphologically segmentable into a detransitivizing prefix and a root \rc{jəpɨ}.

Only the Pemongan languages and \kaxui point to an \rc{j}-initial root, and \kaxui \obj{johɨ} is very rare in contrast to more frequent \obj{o(o)hɨ}.
It only occurs in the third person form of the Progressive, meaning that the \obj{j} may be a reflex of third person \rc{i-}.
As \cref{tab:kaxprog} shows, regular \obj{o}-initial verbs have no third person prefix, while \obj{i-} occurs with C-initial \obj{to[mo]} \qu{to go}.
Even if \qu{to come} once shared this \obj{i-}, the circumstances are strongly in favor of a \obj{j}-initial \kaxui root, allowing the reconstruction of \PC \rc{jəpɨ}.
%
\input{floats/kaxprog}%
%\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
%\a<kax-82>
%\begingl
%\gla johɨ-rɨ//
%\glb \gl{3}.come-\gl{prog}//
%\glft \qu{S/he is coming.}//
%\endgl
%\a<kax-54>
%\begingl
%\gla o-w-ohɨ-rɨ//
%\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
%\glft \qu{You are coming.}//
%\endgl
%\a<kax-77>
%\begingl
%\gla i-nkɨ-rɨ//
%\glb \gl{3}-sleep-\gl{prog}//
%\glft \qu{S/he is sleeping.}//
%\endgl
%\a<kax-83>
%\begingl
%\gla Ø-osone-rɨ//
%\glb \gl{3}-dream-\gl{prog}//
%\glft \qu{S/he is dreaming.}//
%\endgl
%\xe
%
Most of the (non-Pemongan) morphologically complex forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} show no segmental trace of \rc{j}, but the \obj{i} in \akawaio \obj{əsipɨ} is likely a reflex of the sequence \rc{jə}.
This analysis is supported by the reflex \obj{ipɨ} from bare \rc{jəpɨ} in very closely related \macushi.

As for the many forms seemingly reflecting \rc{əpɨ} and \rc{epɨ} rather than \obj{jəpɨ}, they are distributed widely in the family, sometimes even co-occurring in the same language.
A unifying account requires the root \rc{jəpɨ} to undergo two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
While both sound changes are found in other contexts throughout the family \parencite{meira2010origin}, they appear to have applied irregularly to this verb, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as a coalescence of \rc{e} and \rc{əe}.

@@

Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
The scenario suggested here involves reflexes of the detransitivizer \rc{ət(e)-} being optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.
As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ}/\rc{jəpɨ} (\akuriyo) resisted the introduction of \PTir \rc{t-}.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.

>>>>

\subsection{Conservative verbs in comparison}
\label{sec:verbs}
In \cref{sec:extensions}, six incomplete extensions of personal prefixes into \gl{1}\gl{s_a_} territory and the verbs unaffected by them were identified.
This set is rather small in most cases, and many of the verbs remain conservative in different (proto-)languages.
Here, these verbs are investigated from a comparative perspective:
\rc{ka[ti]} \qu{to say} \pcref{sec:say}, \rc{ɨtə[mə]} \qu{to go} \pcref{sec:go}, both roots of the copula \rc{eti}/\obj{a[p]} \pcref{sec:be}, \rc{(ət)jəpɨ} \qu{to come} \pcref{sec:come},  \rc{ɨpɨtə} \qu{to go down} \pcref{sec:godown}, and \rc{e-pɨ} \qu{to bathe} \pcref{sec:bathe}.
The large and phonologically coherent group of \obj{e}-initial verbs not affected by the extension of \akuriyo \obj{k-} \pcref{sec:akuriyo} will not be discussed.

\subsubsection{\rc{ka[ti]} \qu{to say}}
\label{sec:say}
This verb was not affected by any of the extensions in \PPek, \PWai, \PTir, \akuriyo, or \carijo (\crefrange{sec:pekodian}{sec:carijo}), while the first person form of its \yukpa reflex \obj{ka} is unattested.
Most reflexes are simply \obj{ka}, but a fleeting syllable \rc{ti} is reconstructed by \textcite{gildea2007greenberg}, best visible in the imperative forms of some languages. % \exref{kati}.
%
%\ex<kati> \apalai\\
%\obj{kaʃi-ko} \qu{say!}\\
%\parencite[35]{koehn1986apalai}
%\xe
%
\cref{tab:say} shows a comparison of the longest attested forms for each language.%
\footnote{Cognate segments in \crefrange{tab:say}{tab:bathe} were aligned automatically with LingPy \parencite{lingpy268}, for improved exposition of correspondences.}
%
\input{floats/say.tex}%
%

As mentioned in \cref{sec:waiwaian}, \textcite{hixkaryanaderby1985} analyzes the \hixka reflex as transitive, an analytical choice not only motivated by a desire to avoid an idiosyncratic intransitive first person prefix \obj{ɨ-} instead of regular \obj{kɨ-}.
\hixka \obj{ka} also shows the complementary distribution of third person  \obj{n-} and preceding objects typical of transitive verbs in Cariban \parencite[60--81]{gildea1998}.
Due to its semantics, these objects are either ideophones or direct speech \exref{hixsay}.

\pex<hixsay>\hixka
\a<hix-119>
\begingl
\glpreamble onɨ wyaro nkekonɨ bɨryekomo, tɨyonɨ wya//
\gla onɨ wjaro n-ka-jakonɨ bɨrʲekomo tɨ-jonɨ wja//
\glb this like \gl{3}-say-\gl{rem}.\gl{cont} boy \gl{cor}-mother \gl{obl}//
\glft \qu{This is what the boy said to his mother.} \parencite[][36]{hixkaryanaderby1985}//
\endgl
\a<ordpl-35>
\begingl
\glpreamble moro ha, ketxkoná hatá.//
\gla moro ha ka-jat͡ʃkonɨ hatɨ//
\glb \gl{med}.\gl{dem}.\gl{inan} \gl{ints} say-\gl{rem}.\gl{cont}.\gl{pl} \gl{hsy}//
\glft \qu{“That one there” they said.} \parencite[][14]{derbyshire1965textos}//
\endgl
\xe
In \exref{hixsay.hix-119}, the prefix \obj{n-} occurs because there is no preceding object (\qu{he said it like this}), while does not occur in \exref{hixsay.ordpl-35} where the verb is preceded by direct speech.
Looking beyond \hixka, at least the \trio reflex shows the same pattern, albeit inconsistently so \parencite[267]{triocarlin2004}.

\rc{ka[ti]} \qu{to say} also shows transitive patterns in its derivational suffixes:
In \trio, it is the only intransitive verb taking transitive \obj{-po} (\gl{caus}) and the \obj{-ne} (\gl{agt}.\gl{nmlz}) \parencite[263, 169]{triomeira1999}.
It also exceptionally takes the former suffix in \kalina \parencite[82]{courtz2008carib} and \wayana \parencite[258]{wayanatavares2005}.
The agent nominalizer \rc{-ne} became the \panare inflectional suffix \obj{-ɲe} on transitive verbs \parencite[184--185]{gildea1998}.
The combination of \obj{ka} and \obj{-ɲe} likely led \textcite[214]{panarepayne2013} to categorize it as transitive, although \textcite[102]{mattei1994diccionario} arrived at the opposite conclusion.
Finally, reflexes of the \parentext{\parencite{gildea2015valency}} causativizer \rc{-metipo} are found with \obj{ka} in \apalai \parencite[51]{koehn1986apalai} and \waiwai \parencite[52]{waiwaihawkins1998}.

Arugments in favor of intransitive \qu{to say} primarily come from its inflectional prefixes.
\kalina has a minimal pair between transitive \obj{ka} \qu{to remove} and intransitive \obj{ka} \qu{to say}, \obj{sikai} \qu{I took it away} vs \obj{wɨkai} \qu{I said} \parencite[288, 45]{courtz2008carib}.
Similarly, \PPek \rc{ke} \qu{to say} took \gl{1}\gl{s_a_} \rc{w-} \pcref{sec:pekodian}, rather than \gl{1}>\gl{3} \obj{s-} (\bakairi) or \rc{ini-} (\PXin).
Additionally, languages differentiating transitive from \gl{s_a_} prefixes by the presence of \obj{i} \parencite[495]{meira2010origin} have \obj{ɨ}-final prefixes, see \akuriyo in \exref{sayintr.aku-154}, as well as \textcites[294]{triomeira1999}[195]{wayanatavares2005}[288]{ikpengpacheco2001}[150]{alves2017arara}[168]{hoff1968carib} for cognate forms in other such languages.
Finally, the \gl{s_a_} class marker \obj{w-} occurs on nominalizations in \kalina \exref{sayintr.kar-84}, and is probably reflected as vowel length in the \trio \parencite[333]{triomeira1999} and \wayana \parencite[196]{wayanatavares2005} participles.

@@

\subsubsection{\rc{ɨtə[mə]} \qu{to go}}
\label{sec:go}
This verb was not affected by the any of the extensions in \cref{sec:extensions}.
\textcite{gildea2007greenberg} reconstruct it as \rc{tə[mə]}, like \rc{ka[ti]} \qu{to say} with a fleeting second syllable.
While many reflexes are clearly \obj{t}-initial \parentext{e.g. \hixka \obj{ntoje} \qu{he went} \parencite[27]{hixkaryanaderby1985}, \trio \obj{təkə} \qu{go!} \parencite[246]{triomeira1999}, or \wayana \obj{kuptəm} \qu{we went} \parencite[195]{wayanatavares2005}}, an initial vowel \rc{ɨ} must clearly be reconstructed \pcref{tab:go}%
\footnote{The brackets in \cref{tab:go} show that languages have the initial vowel only in some forms.
Also, the prefix-stem boundary in many inflected forms like \trio \obj{wɨtənne} or \arara \obj{wɨdolɨ} \qu{I went} \parencites[43]{triomeira1999}[153]{alves2017arara} is ambiguous, since epenthetic \obj{ɨ} breaks up CC clusters on these boundaries.
Still, when considering unambiguous forms, the contrast with C-initial \rc{ka[ti]} becomes very clear.}

@@

\subsubsection{\rc{eti} and \rc{a[p]} \qu{to be}}
\label{sec:be}
\rc{a[p]} is the older copula and already had various irregularities in \PC \parencite{gildea2018reconstructing}.
\rc{eti} is reconstructed by \textcites{meira2009property}{gildea2018reconstructing} as originally meaning \qu{to dwell, live}, but serving as a copula already in \PC.\footnote{Such a stative, locative source is also suggested by the existence of \obj{it͡ʃi} \qu{to lie down} in \arara \parencite[196]{alves2017arara}.}
Modern reflexes of these roots are used suppletively, conditioned by person and\slash{}or \gl{tam}.
Both roots preserved \gl{1}\gl{s_a_} \rc{w-} in \PPek, \PWai, and \PTir (\crefrange{sec:pekodian}{sec:taranoan}).
\akuriyo \obj{a} was not affected by the extension of \obj{k-} \pcref{sec:akuriyo}, while the first person form of \obj{eʔi} is not attested. 
\carijo innovated \obj{j-}, but only in the reflex of \rc{eti} \exref{mivida-12}; the \obj{a} root preserves \obj{w-} \pcref{sec:carijo}.
\yukpa introduced \obj{j-} to the reflexes of both \rc{a[p]} and \rc{eti}, which are preserved as encliticized auxiliaries in certain constructions \exref{yukcop}.

@@

\ex<yukcop> \yukpa \parencite[143--144]{meira2006syntactic}\\
\begin{tabular}[t]{@{}lll@{}}
	& \gl{npst} & \gl{pst}\\
	\gl{1} & \obj{=j-a(-s)}&\obj{=j-e}\\
	\gl{2} & \obj{=mak(o)}&\obj{=m-e}\\
	\gl{3} & \obj{=mak(o)}&\obj{=n-e}\\
\end{tabular}
\xe
%
A comprehensive comparative overview of these two roots is given by \textcite[375--382]{gildea2018reconstructing}; they will not be discussed in detail here.

\subsubsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}
This verb can be reconstructed as \PC \rc{(ət-)jəpɨ}, morphologically segmentable into an optional detransitivizing prefix and a root \rc{jəpɨ}.
As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ}/\rc{jəpɨ} (\akuriyo) resisted the introduction of \PTir \rc{t-}.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
No \yukpa reflex of this verb is attested, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come}.

@@

This reconstruction disagrees with \posscite[30]{gildea2007greenberg} \rc{ətepɨ}, due to the comparative evidence in \cref{tab:come}.
Crucially, most reflexes do not reflect the \rc{ət} part of \posscite{gildea2007greenberg} reconstruction, and many ostensibly go back to \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ}, rather than \rc{epɨ}.

Only the Pemongan languages and \kaxui point to an \rc{j}-initial root, and \kaxui \obj{johɨ} is very rare in contrast to more frequent \obj{o(o)hɨ}.
It only occurs in the third person form of the Progressive, meaning that the \obj{j} may be a reflex of third person \rc{i-}.
However, regular \obj{o}-initial \kaxui verbs have no third person prefix \pcref{tab:kaxprog}, while \obj{i-} occurs with C-initial \obj{to[mo]} \qu{to go}.
Even if \qu{to come} once shared this \obj{i-}, the \obj{j} is best analyzed as part of the root in \kaxui, allowing the reconstruction of \PC \rc{jəpɨ}.
%
\input{floats/kaxprog}%
%\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
%\a<kax-82>
%\begingl
%\gla johɨ-rɨ//
%\glb \gl{3}.come-\gl{prog}//
%\glft \qu{S/he is coming.}//
%\endgl
%\a<kax-54>
%\begingl
%\gla o-w-ohɨ-rɨ//
%\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
%\glft \qu{You are coming.}//
%\endgl
%\a<kax-77>
%\begingl
%\gla i-nkɨ-rɨ//
%\glb \gl{3}-sleep-\gl{prog}//
%\glft \qu{S/he is sleeping.}//
%\endgl
%\a<kax-83>
%\begingl
%\gla Ø-osone-rɨ//
%\glb \gl{3}-dream-\gl{prog}//
%\glft \qu{S/he is dreaming.}//
%\endgl
%\xe
%
The morphologically complex (non-Pemongan) forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} show no segmental trace of \rc{j}, but the \obj{i} in \akawaio \obj{əsipɨ} likely reflects the sequence \rc{jə}.
This analysis is supported by bare \rc{jəpɨ} becoming \obj{ipɨ} in closely related \macushi.

The many forms seemingly reflecting \rc{əpɨ} and \rc{epɨ} rather than \obj{jəpɨ} are distributed unsystematically in the family, sometimes even co-occurring in the same language.
A unifying account requires the root \rc{jəpɨ} to undergo two main sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
Both sound changes are found in other contexts throughout the family \parencite{meira2010origin}, but they appear to have applied inconsistently to this verb, and not always in the same order.
Forms like \kalina \obj{opɨ} can only be explained if \rc{j} was lost before the umlauting of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as a coalescence of \rc{e} and \rc{əe}.

@@

Summing up, this verb is highly irregular, both from a synchronic and diachronic perspective.
The scenario suggested here involves reflexes of the detransitivizer \rc{ət(e)-} being optionally added to an \gl{s_a_} verb root \rc{jəpɨ}, which further underwent umlaut and loss of \rc{j}, but in no systematic manner, resulting in the chaotic picture in \cref{tab:come}.
